---
title: "STACCato Decomposition of SLE dataset"
author: "Qile Dai"
date: "2023-11-26"
output: 
  rmarkdown::html_document:
    number_sections: true
    toc: true
    theme: united
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Settings

```{r message=FALSE, warning=FALSE}
########## Setting ###################
set.seed(2018)
library(tensorregress)
library(R.matlab)
library(ggplot2)
library(dplyr)
library(parallel)
library(reticulate)
library(reshape2)
library(gridExtra)

source('https://raw.githubusercontent.com/daiqile96/STACCato/main/R/dcomp.R')
data.dir = 'Data'
```

# Load Data

Read in sample-level covariates:

```{r}
# read in covariates data
cov.mat <- read.table(file.path(data.dir, 'sample_info.txt'), 
                      header = T, sep = '\t') %>% as.data.frame() %>% 
  mutate(pop_cov_European = ifelse(pop_cov == 'European', 1, 0),
         Sex_Male = ifelse(Sex == 'Male', 1, 0),
         Processing_Cohort_2 = ifelse(Processing_Cohort == 2, 1, 0),
         Processing_Cohort_3 = ifelse(Processing_Cohort == 3, 1, 0),
         Processing_Cohort_4 = ifelse(Processing_Cohort == 4, 1, 0),
         Age = scale(Age)[,1],
         SLE_status_SLE = ifelse(SLE_status == 'SLE', 1, 0))
```

Create sample-level design matrix:

```{r}
cov.ind <- as.matrix(cbind('intercept' = rep(1, nrow(cov.mat)), 
                           cov.mat[, c('Age', 'Sex_Male', 
                                       'pop_cov_European',
                                       'Processing_Cohort_2',
                                       'Processing_Cohort_3',
                                       'Processing_Cohort_4',
                                       'SLE_status_SLE')]))
cov.names = colnames(cov.ind)
```

Read in cell cell communication score tensor:

```{r}
mat <- readMat(file.path(data.dir, 'c2ctensor.mat'))
c2ctensor <- as.tensor(mat$c2ctensor)
```

Read in cell type names:

```{r}
lrs = read.table(file.path(data.dir, "lr.txt"), 
                 header = F,
                 sep = '\t') %>% unlist()
senders = read.table(file.path(data.dir, "sender.txt"), 
                     header = F,
                     sep = '\t')  %>% unlist()
receivers = read.table(file.path(data.dir, "receiver.txt"), 
                       header = F,
                       sep = '\t')  %>% unlist()
```

# Decomposition 
## Determine decomposition rank

Determine the number of components for sender cell type, receiver cell type, and ligand-receptor pair dimension.

```{r}
# Determine the number of components for sender cell type
r2 = cal_eigen_varexp(c2ctensor, 2, c(1,3,4), varexp = 1)
# Determine the number of components for receiver cell type
r3 = cal_eigen_varexp(c2ctensor, 3, c(1,2,4), varexp = 1)
# Determine the number of components for ligand-receptor pair
r4 = cal_eigen_varexp(c2ctensor, 4, c(1,2,3), varexp = 1)

plot(r2$p)

plot(r3$p)

plot(r4$p)
```

## Perform supervised tensor decomposition 
```{r results='hide'}
# decomposition rank
r = c(ncol(cov.ind), r2$rank, r3$rank, r4$rank)
# staccato decomposition
staccato_all_cov = staccato(tsr=c2ctensor, 
                            X_covar1=cov.ind,
                            lr.names = lrs,
                            sender.names = senders,
                            receiver.names = receivers,
                            core_shape = r)
```

# Estimated effects of disease on Communication Events

## Extract estimated SLE disease effects

```{r}
disease_eff = get_full_effect(staccato_all_cov$C_ts, 
                              var = 'SLE_status_SLE',
                              val = 'effect')
```

## Assess significance level of estimated SLE disease effects

We performed parametric bootstrap to get p-values for estimated effects.

```{r}
staccato_all_cov_P = boot_p(n_boot = 999,
                            dcomp_res = staccato_all_cov,
                            n_thread = 6,
                            lr.names = lrs,
                            sender.names = senders,
                            receiver.names = receivers)

disease_pval = get_full_effect(staccato_all_cov_P, 
                               var='SLE_status_SLE',
                               val = 'pval')
head(disease_pval)

disease_eff_pval = merge(disease_eff, disease_pval,
                         by = c('LR', 'Sender', 'Receiver'))
```

## Determine significant estimated SLE disease effects

```{r fig.width=6}
disease_eff_pval = merge(disease_eff, disease_pval,
                         by = c('LR', 'Sender', 'Receiver')) 
plot_eff_pval_sig(disease_eff_pval,
                            eff_cutoff = 0.015,
                            pval_cutoff = 5e-02) +
  theme(text = element_text(face='bold'))
```

```{r}
disease_eff_sig = get_eff_sig(disease_eff_pval, 0.015, 5e-02)  
cell_level = c('B', 'NK', 'T4', 'T8', 'Prolif', 'cDC', 'pDC','cM', 'ncM')
cell_label = c('B', 'NK', 'CD4+ T', 'CD8+ T', 'Prolif', 'cDC', 'pDC', 'cM', 'ncM')
disease_eff_sig = disease_eff_sig %>% 
  mutate(Sender = factor(Sender, levels = cell_level, labels = cell_label),
         Receiver = factor(Receiver, levels = cell_level, labels = cell_label))
```


## Plot significant SLE disease effect by cell types

```{r fig.width=12, fig.height=8}
eff_df = disease_eff_sig %>% 
  filter(Receiver %in% c('B', 'CD8+ T', 'pDC', 'cM')) %>% 
  select(LR, Sender, Receiver, effect) %>% 
  mutate(Receiver = paste0('Receiver: ', Receiver))

plot_disease_eff_by_receiver_cell(eff_df) +
  theme(text = element_text(face='bold'))
```


# Check activity patterns of cell types

## Plot cell factors

```{r}
p_factors = plot_sender_receiver(dcomp_res = staccato_all_cov,
                                 cell_level = cell_level,
                                 cell_label = cell_label) 
grid.newpage()
grid.draw(p_factors)
```

## Evaluate and plot contributions of cell factors

```{r fig.width=8, fig.height=4}
p_sender = get_contribution_by_mode(mode = 'W3',
                                    dcomp_res = staccato_all_cov) +
  theme(text = element_text(face='bold'))

p_receiver = get_contribution_by_mode(mode = 'W4',
                                      dcomp_res = staccato_all_cov) +
  theme(text = element_text(face='bold'))

p_contributions = gridExtra::arrangeGrob(grobs = list(p_sender,
                                         p_receiver), 
                                         ncol = 2)
grid.newpage()
grid.draw(p_contributions)
```



