---
title: "Apply STACCato to identify condition-related CCC events"
author: "Qile Dai"
date: "2024-10-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Inputs

The inputs for this example can be downloaded from [here](https://www.dropbox.com/scl/fo/ffk3dm4y8fgklqtv5srqw/ALAH3njDav9xNeY0BmliOBU?rlkey=cb9ubvcsyejcvv6bpr8c5h2u7&st=7lnap5zv&dl=0). I saved the inputs under the `data.dir`. 

```{r}
data.dir = '~/projects/Projects/STDCC/Data/RDA/ASD/LIANA_expr_prod'
```


# Settings

```{r message=FALSE, warning=FALSE}
########## Setting ###################
set.seed(2018)
library(tensorregress)
library(R.matlab)
library(ggplot2)
library(dplyr)
library(parallel)
library(reticulate)
library(reshape2)
library(gridExtra)

source('https://raw.githubusercontent.com/daiqile96/STACCato/main/R/dcomp.R')
```

# Load Data

Read in individual-level covariates:

```{r}
# read in covariates data
cov.mat <- read.table(file.path(data.dir, 'subjects_info.txt'),
                      header = T, sep = '\t') %>% as.data.frame() %>%
  mutate(sex_M = ifelse(sex == 'M', 1, 0),
         Seqbatch_SB1 = ifelse(Seqbatch == 'SB1', 1, 0),
         age = scale(age)[,1],
         diagnosis_ASD = ifelse(diagnosis == 'ASD', 1, 0))
```

Create individual-level design matrix:

```{r}
cov.ind <- as.matrix(cbind('intercept' = rep(1, nrow(cov.mat)),
                           cov.mat[, c('age', 'sex_M',
                                       'Seqbatch_SB1',
                                       'diagnosis_ASD')]))
cov.names = colnames(cov.ind)
```

Read in cell cell communication score tensor:

```{r}
mat <- readMat(file.path(data.dir, 'c2ctensor.mat'))
c2ctensor <- as.tensor(mat$c2ctensor)
```

- check dimension
```{r}
dim(c2ctensor@data)
```

- check zero proportions

```{r}
1 - mean(c2ctensor@data != 0)
```

Read in cell type names and ligand-receptor pairs:

```{r}
lrs = read.table(file.path(data.dir, "lr.txt"),
                 header = F,
                 sep = '\t') %>% unlist()
senders = read.table(file.path(data.dir, "sender.txt"),
                     header = F,
                     sep = '\t')  %>% unlist()
receivers = read.table(file.path(data.dir, "receiver.txt"),
                       header = F,
                       sep = '\t')  %>% unlist()
```


# Decomposition
## Determine decomposition rank

Determine the number of components for sender cell type, receiver cell type, and ligand-receptor pair dimension.

```{r}
# Determine the number of components for sender cell type
r2 = cal_eigen_varexp(c2ctensor, 2, c(1,3,4), varexp = 1)
# Determine the number of components for receiver cell type
r3 = cal_eigen_varexp(c2ctensor, 3, c(1,2,4), varexp = 1)
# Determine the number of components for ligand-receptor pair
r4 = cal_eigen_varexp(c2ctensor, 4, c(1,2,3), varexp = 1)

plot(r2$p)

plot(r3$p)

plot(r4$p)
```

## Perform tensor-based regression

```{r results='hide'}
# decomposition rank
r = c(ncol(cov.ind), r2$rank, r3$rank, r4$rank)
# staccato
staccato_all_cov = staccato(tsr=c2ctensor,
                            X_covar1=cov.ind,
                            lr.names = lrs,
                            sender.names = senders,
                            receiver.names = receivers,
                            core_shape = r)
```

## Top 1000 Communication Events

```{r, fig.width=26, fig.height=28}
disease_eff = get_full_effect(staccato_all_cov$C_ts,
                              var = 'diagnosis_ASD',
                              val = 'effect')

select_pattern = disease_eff %>%
  mutate(Pattern = paste(LR, Sender, Receiver, sep = ' - ')) %>%
  arrange(desc(abs(effect))) %>%
  slice(1:1000) %>%
  dplyr::select(Pattern) %>%
  unlist(.) %>% as.character(.)

disease_eff_top = disease_eff %>%
  mutate(Pattern = paste(LR, Sender, Receiver, sep = ' - ')) %>%
  filter(Pattern %in% select_pattern)

# top_cells = unique(disease_eff_pval$Sender)[8:17]

disease_eff_top$panel <- ifelse(disease_eff_top$Receiver %in% c("AST-FB", "AST-PP", "Endothelial",
                                                                "IN-PV", "IN-SST", "IN-SV2C", "IN-VIP"),
                                "Panel 1", "Panel 2")

max = max(abs(disease_eff_top$effect))

p_compare_receiver_selected = ggplot(disease_eff_top %>%
                                       mutate(Receiver = paste('Receiver:', Receiver)),
                                     aes(Sender, LR)) +    # Create default ggplot2 heatmap
  geom_tile(aes(fill = effect)) +
  scale_fill_gradient2(name = 'Disease Effect',
                       limits = c(-1 * max, max),
                       low = "blue",
                       mid = "white",
                       high = "red",
                       guide = guide_colorbar(barwidth = 30)) +
  theme_linedraw() +
  theme(text = element_text(size = 20, face = 'bold'),
        axis.text.x = element_text(angle=90,vjust = 1, hjust=1),
        legend.position = 'top',
        legend.text = element_text(size = 16),
        strip.background = element_rect(fill = "white"),
        strip.text = element_text(color = 'black')) +
  facet_wrap(~Receiver, ncol = 6) +
  xlab('Sender Cell Types') +
  ylab('Ligand - Receptor Pairs')

p_compare_receiver_selected
```

## Perform Bootstrap to get p-values for estimated effects

While fitting of a single STACCato tensor-based regression only takes seconds, assessing the significance level of estimated effects by bootstrapping (requiring multiple iterations of model fitting) takes hours of CPU time. Here we performed 20 times boostrap as an example to show how to use bootstrap to get p-values.

```{r}
# bootstrap
n_boot = 20
n_thread = 5
P = boot_p(n_boot = n_boot,
           dcomp_res = staccato_all_cov,
           n_thread = n_thread,
           lr.names = lrs,
           sender.names = senders,
           receiver.names = receivers)
```


```{r}
disease_pval = get_full_effect(P,
                               var='diagnosis_ASD',
                               val = 'pval')
head(disease_pval)
```

